Name: Prosper Loan dataset analysis
========================================================
```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
#load packages
library(ggplot2)
library(dplyr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data

loandata = read.csv('~/Downloads/prosperLoanData.csv')
#Display 6 rows to review the data in general as a quick view then 
#display the 81 vars in details in order to help writing 
#a short introduction and perform EDA

head(loandata)

```


This dataset contains 113,937 loans with 81 variables; EDA will be perofromed 
for some variables as follows: Univariate Plots, Bivariate Plots, 
and Multivariate Plots. 
Prosper is a platform which is a good option for those who can’t get a loan 
from a traditional bank and don’t want the high interest rates offered by 
credit cards and payday loans. the process workflow based on the actors is:
borrower: submit a loan application. 
prosper: provide loans after doing checks with some orgnizations to make sure 
the borrower meets several criteria. 


# Univariate Plots Section


```{r echo=FALSE,warning=FALSE, message=FALSE, Univariate99}
#First, i will do Univariate anaylsis which performs for one variable 
#as an exploration to understand the data patterns and its distrbuation
#for each plot i will write a result and in the section below 
#i will conclude them:

#ploting LoanOriginalAmount for checking the loans amount
ggplot(loandata,aes(x=LoanOriginalAmount))+
geom_histogram(binwidth=1000)+
scale_x_continuous(breaks=seq(0,36000,2000),lim=c(0,30000)) +  
  theme(axis.text.x = element_text(angle=90, hjust=1))
summary(loandata$LoanOriginalAmount)
#as a result, from the histogram above the most loan amount is 4000 then 10000 
#and 15000. so the amount peaks is between $4000 and $14000 
```

from the histogram above the most loan amount is $4000 then $10000 
and $15000. so the amount peaks is between $4000 and $14000 

```{r echo=FALSE,warning=FALSE, message=FALSE, Univariate2}
#ploting BorrowerRate to check how much the borrower assured monthly payments
ggplot(aes(BorrowerRate), data=loandata) + 
geom_histogram()
summary(loandata$BorrowerRate)
 
 
```

from th histogram above the  borrower rate is around 20% 
and the range is between 0.1 and 0.3 (10% to 30%)


```{r echo=FALSE,warning=FALSE, message=FALSE, Univariate3}
#ploting ProsperScore to check the probability of a loan going bad
ggplot(aes(ProsperScore),data=loandata)+
geom_bar()
summary(loandata$ProsperScore)
#Borrowers proposer's scores are between 1.00 and 11.00 with median 6.00 
```

from the bar chart Borrowers proposer's scores are between 1.00 and 11.00 with median 6.00 

```{r echo=FALSE,warning=FALSE, message=FALSE, Univariate}
#ploting how many loans per each loanStatus
#reorder loan status

table(loandata$LoanStatus)
loandata$LoanStatus<-ordered(loandata$LoanStatus,
                             levels=c('Current','Completed', ' Chargedoff', 
                                      'FinalPaymentInProgress', 'Defaulted',
                                      'Cancelled',
                                      'Past Due (1-15 days)',
                                      'Past Due (16-30 days)',
                                      'Past Due (31-60 days)',
                                      'Past Due (61-90 days)',
                                      'Past Due (91-120 days)',
                                      'Past Due (>120 days)'))



#ploting loanstatus
ggplot(aes(LoanStatus),data=loandata)+geom_bar() +
  theme(axis.text.x = element_text(angle=90, hjust=1))
summary(loandata$LoanStatus)
```

from the bar chart, it is clear that most loan data is for current loans 
then, completed loans and then chargedoff loans. we have some of cases 
which are paid after the due date.

```{r echo=FALSE,warning=FALSE, message=FALSE, Univariate4}
#ploting EmploymentStatus for the borrowers to check what is the most employment
#status for borrowers
#reorder EmploymentStatus
loandata$EmploymentStatus<-ordered(loandata$EmploymentStatus,levels=c('Employed',
                                    'Full-time', 'Part-time', 'Self-employed',
                                    'Retired', 'Not employed','Not available',
                                    'Other','NA'))
ggplot(aes(EmploymentStatus),data=loandata)+geom_bar() +
  theme(axis.text.x = element_text(angle=90, hjust=1))
  summary(loandata$EmploymentStatus)
```

from the bar chart above the most of borrowers employment status 
is employed then full-time 

```{r echo=FALSE,warning=FALSE, message=FALSE, Univariate5}
#ploting StatedMonthlyIncome for borrowers to check for borrowers 
  #theier Monthly Income
ggplot(aes(StatedMonthlyIncome),data=loandata) +       
geom_histogram(binwidth=500) + xlim(0,10000)
summary(loandata$StatedMonthlyIncome)
# as a result, from the histogram above borrowers monthly income range 
#is between 25,000$ and 75,000$
```

from the histogram above borrowers monthly income range 
is between 25,000$ and 75,000$

# Univariate Analysis

### What is the structure of your dataset?
This dataset contains 113,937 loans with 81 variables. Prosper is a platform 
which is a good option for those who can’t get a loan from a traditional bank 
and don’t want the high interest rates offered by credit cards and payday loans. 
the process workflow based on the actors is:
#borrower: submit a loan application. 
#prosper: provide loans after doing checks with some orgnizations to 
make sure the borrower meets several criteria. 

### What is/are the main feature(s) of interest in your dataset?
The main features of interest in the dataset are BorrowerRate and ProsperScore. 
from the invstigation above there is a relashionship between them 
as the ProsperScore affect BorrowerRate

### What other features in the dataset do you think will help support your \
Other features: LoanStatus, EmploymentStatus, StatedMonthlyIncome, 
and LoanOriginalAmount might affect BorrowRate.

### Did you create any new variables from existing variables in the dataset?
No

### Of the features you investigated, were there any unusual distributions? \
I reordered LoanStatus and EmploymentStatus as they have unordered data.




# Bivariate Plots Section

```{r echo=FALSE,warning=FALSE, message=FALSE, Bivariate_Plots}
#exlporation for two variables (attributes). 
#It explores the concept of relationship between two variables.


#first, ploting two variables for the relationship between 
#(ProsperScore,BorrowerRate)
#we will check how much the borrower assured monthly payments vs. Prosper Score
ggplot(aes(ProsperScore,BorrowerRate),data=loandata)+
geom_jitter(alpha=0.01)+
  scale_x_continuous(breaks=seq(0,12,1))
cor.test(loandata$BorrowerRate,loandata$ProsperScore)
```

from the chart above we observe that when BorrowerRate decreases 
ProsperScore increases.
correlation coefficient is -0.65 
BorrowerRate has strong relationship with ProsperScore.

```{r echo=FALSE,warning=FALSE, message=FALSE, Bivariate_Plotss}
#ploting relationship between BorrowerRate and StatedMonthlyIncome and
#we will check how much the borrower assured monthly payments vs.  vs. the borrower Monthly Income
ggplot(aes(x = StatedMonthlyIncome, y = BorrowerRate), data=loandata) +
  geom_point(alpha=1/100) +
  geom_smooth(colour = "red", show.legend = NA)
cor.test(loandata$BorrowerRate,loandata$StatedMonthlyIncome)
```

from the chart above we observe tha when BorrowerRate decreases 
StatedMonthlyIncome increases.
correlation coefficient is -0.33, so a BorrowerRate moderate relationship.


```{r echo=FALSE,warning=FALSE, message=FALSE, Bivariate_Plotsss}
#ploting LoanStatus and BorrowerRate 
#we will check how much the borrower assured monthly payments vs each loan status
ggplot(aes(LoanStatus,BorrowerRate),data=loandata)+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

from the chart above we observe the mean of BorrowerRate for current 
and completed are less than late payemnet in gereneral (i.e. pastdue), 
defaulted and chargedoff 

```{r echo=FALSE,warning=FALSE, message=FALSE, Bivariate_Plotss1}
#ploting EmploymentStatus, and BorrowerRate
#we will check how much the borrower assured monthly payment vs each borrower Employment Status
ggplot(aes(EmploymentStatus,BorrowerRate),data=loandata)+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

from the chart above we obeserve that the mean of BorrowerRate for 
not employed borrower is higher than others.

```{r echo=FALSE,warning=FALSE, message=FALSE, Bivariate_Plotss2}
#other interesting features relationship
#ploting LoanStatus, and StatedMonthlyIncome
#we will check for each loan status how much borrower assured monthly payment
ggplot(aes(LoanStatus, StatedMonthlyIncome),data=loandata)+
  geom_boxplot()+
  ylim(0, quantile(loandata$StatedMonthlyIncome,0.99))+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

from the chart above we observe that current loan status is the highest 
StatedMonthlyIncome 
and late paymnet in gereneral(i.e. pastdue)loan has lower StatedMonthlyIncome.

```{r echo=FALSE,warning=FALSE, message=FALSE, Bivariate_Plotss3}
#ploting ProsperScore, BorrowerRate and LoanStatus per each status
#Finding relationship between categorical and continuous variables
#we will check prosper score and how much borrower assured monthly payment 
#for each loan status
 ggplot(aes(ProsperScore, BorrowerRate), 
        data=subset(loandata,loandata$LoanStatus!='Completed'))+
  geom_jitter(alpha=0.01)
``` 

the chart above presents that prosper score and how much borrower 
assured monthly payment for all loan status exculds Completed status

```{r echo=FALSE,warning=FALSE, message=FALSE, Bivariate_Plotss33}
  ggplot(aes(ProsperScore, BorrowerRate), 
         data=subset(loandata,loandata$LoanStatus!='Current'))+
  geom_jitter(alpha=0.01)
```
  
the chart above presents that prosper score and how much borrower 
assured monthly payment for all loan status exculds Current status


```{r echo=FALSE,warning=FALSE, message=FALSE, Bivariate_Plotss3333}
  ggplot(aes(ProsperScore, BorrowerRate), 
         data=subset(loandata,loandata$LoanStatus!='PastDue'))+
  geom_jitter(alpha=0.01)
```

the chart above presents that prosper score and how much borrower 
assured monthly payment for all loan status exculds PastDue status

```{r echo=FALSE,warning=FALSE, message=FALSE, Bivariate_Plotss333333}

    ggplot(aes(ProsperScore, BorrowerRate), 
           data=subset(loandata,loandata$LoanStatus!='Chargedoff'))+
  geom_jitter(alpha=0.01)
```

the chart above presents that prosper score and how much borrower 
assured monthly payment for all loan status exculds Chargedoff status

```{r echo=FALSE,warning=FALSE, message=FALSE, Bivariate_Plotss34333}
    ggplot(aes(ProsperScore, BorrowerRate), 
           data=subset(loandata,loandata$LoanStatus!='Defaulted'))+
  geom_jitter(alpha=0.01)
```

the chart above presents that prosper score and how much borrower 
assured monthly payment for all loan status exculds Defaulted status




# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
we observe the following:

the mean of BorrowerRate decreases while ProsperScore increases. 
correlation coefficient is -0.65 
so BorrowerRate has strong relationship with ProsperScore. 

the mean of BorrowerRate decreases while LoanOriginalAmount increases. 
correlation coefficient is -0.33, so BorrowerRate has moderate relationship
with LoanOriginalAmount

the mean of BorrowerRate decreases while StatedMonthlyIncome increases. 
correlation coefficient is -0.088, so BorrowerRate has weak relationship.
with StatedMonthlyIncome

the mean of BorrowerRate for current and completed loan status are less than 
late payment in gereneral (i.e. pastdue), defaulted and chargedoff. 
the mean of BorrowerRate for not employed borrower is higher than others.

### Did you observe any interesting relationships between the other features?
relationships between LoanStatus, and StatedMonthlyIncome so:
the current loan status is the highest StatedMonthlyIncome and late payemnet 
in gereneral (i.e. pastdue) loan has lower StatedMonthlyIncome.

### What was the strongest relationship you found?
the strongest relationship is between BorrowerRate and ProsperScore 
since when the mean of BorrowerRate decreases the ProsperScore increases. 
as well,between BorrowerRate and LoanOriginalAmount moderate relationship. 
so, BorrowerRate will be affaected by ProsperScore and LoanOriginalAmount. 



# Multivariate Plots Section




```{r echo=FALSE,warning=FALSE, message=FALSE, Multivariatee}
#relashionship between three variables ProsperScore, BorrowerRate and LoanStatus 
ggplot(aes(ProsperScore, BorrowerRate),
       data=loandata) + 
  geom_jitter(aes(color=LoanStatus),alpha=0.01, show.legend = FALSE)+ 
  facet_wrap(~LoanStatus)+
   ggtitle('Facet Wrap by LoanStatus')
 
```

from the chart above there are a lot of current and completed loans 
with lower BorrowerRate and higher ProsperScore

```{r echo=FALSE,warning=FALSE, message=FALSE, Multivariatee1}
ggplot(aes(x=ProsperScore, y=BorrowerRate),
               data = subset(loandata,!is.na(LoanStatus))) +
    geom_point(stat="summary",aes(color=LoanStatus),show.legend = FALSE,
             fun.y = mean, size = 1) +
    ylab('Mean BorrowerRate') +
    facet_wrap(~LoanStatus)+
  geom_smooth(method=lm)+
    ggtitle('Mean ProsperScore vs BorrowerRate by LoanStatus')+
    theme(plot.title=element_text(hjust=0.5))

```  


from the chart above we confirm that the mean of BorrowerRate decreases 
while ProsperScore increases for loan status such as current and completed.




```{r echo=FALSE,warning=FALSE, message=FALSE, Multivariatee2}
ggplot(aes(x=ProsperScore, y=BorrowerRate),
               data = subset(loandata,!is.na(EmploymentStatus))) +
    geom_point(stat="summary",aes(color=EmploymentStatus),show.legend = FALSE,
             fun.y = mean, size = 1) +
    ylab('Mean BorrowerRate') +
   geom_smooth(method=lm)+
    facet_wrap(~EmploymentStatus)+
    ggtitle('Mean ProsperScore vs BorrowerRate by EmploymentStatus')+
    theme(axis.text.x = element_text(angle = 0, hjust = 1))
```  


from the chart above we confirm that the mean of BorrowerRate decreases 
while ProsperScore increases for employment status.


# Multivariate Analysis

  
### Talk about some of the relationships you observed in this part of the \
with lower BorrowerRate we will have higher ProsperScore and 
this is confirmed with some of LoanStatus such as completed and current.


### Were there any interesting or surprising interactions between features?
interesting relationship between BorrowerRate,ProsperScore and LoanStatus were confirmed above.

------

# Final Plots and Summary


### Plot One
```{r echo=FALSE, warning=FALSE, message=FALSE, Plot_One}
#ploting BorrowerRate 
#ploting BorrowerRate to check how much the borrower assured monthly payments
ggplot(aes(BorrowerRate), data=loandata) + 
geom_histogram(color='black')+
    ggtitle('BorrowerRate')

```

### Description One
The plot above describes BorrowRate for NumberOfLoans from ~ .05 to ~ .35 
the peak is at ~ 0.15.

### Plot Two
```{r echo=FALSE, warning=FALSE, message=FALSE,  Plot_Two}
#ploting ProsperScore, and BorrowerRate to check how much the borrower assured monthly payments vs. Prosper Score
ggplot(aes(ProsperScore,BorrowerRate),data=loandata)+
geom_jitter(alpha=0.01)+
   ggtitle('BorrowerRate by ProsperScore')+
  scale_x_continuous(breaks=seq(0,12,1))
cor.test(loandata$BorrowerRate,loandata$ProsperScore)
```

### Description Two
when BorrowerRate decreases ProsperScore increases. 
correlation coefficient is -0.65 so BorrowerRate has strong 
relationship with ProsperScore. 

### Plot Three
```{r echo=FALSE,warning=FALSE, message=FALSE, Plot_Three}
#relashionship between three variables 

ggplot(aes(x=ProsperScore, y=BorrowerRate),
               data = subset(loandata,!is.na(LoanStatus))) +
    geom_point(stat="summary",aes(color=LoanStatus),show.legend = FALSE,
             fun.y = mean, size = 1) +
    ylab('Mean BorrowerRate') +
    facet_wrap(~LoanStatus)+
    geom_smooth(method=lm)+
    ggtitle('Mean ProsperScore vs BorrowerRate by LoanStatus')+
    theme(plot.title=element_text(hjust=0.5))

```  


  

### Description Three

the mean of BorrowerRate decreases while ProsperScore increases for 
loan status such as current and completed.
------

#Reflection
This dataset contains 113,937 loans with 81 variables; EDA has been perofromed 
for some variables such as BorrowerRate, ProsperScore, , LoanOriginalAmount, 
LoanStatus, StatedMonthlyIncome and EmploymentStatus.
1st, Univariate Plots for one variable.
then, Bivariate Plots for two variables.
Finally, Multivariate Plots for categorical and continuous variables
I foucsed on BorrowerRate and its relationship with other variables.
we found the following:
the mean of BorrowerRate decreases while ProsperScore increases. 
BorrowerRate has strong relationship with ProsperScore. 

the mean of BorrowerRate decreases while LoanOriginalAmount increases. 
BorrowerRate has moderate relationship with LoanOriginalAmount

the mean of BorrowerRate decreases while StatedMonthlyIncome increases. 
BorrowerRate has weak relationship with StatedMonthlyIncome

the mean of BorrowerRate for current and completed loan status are less than 
late payment in gereneral (i.e. pastdue), defaulted and chargedoff. 
the mean of BorrowerRate for not employed borrower is higher than others.

with lower BorrowerRate we will have higher ProsperScore and 
this is confirmed with some of LoanStatus such as completed and 
current.

for future work, I may explore more variables periodically to improve the process workflow



